SSA based global code motion of identical computations

### 2017-04-16 acmart-pacmpl-template.tex
refactoring

\section{Introduction}
Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like CSE \cite{dragonbook} work very well on single basic blocks but
when it comes to detect redundancies across basic blocks these techniques fall
short: more complex passes like GCSE and \PRE{} have been designed to handle these
cases based on dataflow analysis \cite{morel1979global}.  At first these
techniques were described in the classical data-flow analysis framework, and
later the use of the \SSA{} representation lowered the cost in terms of compilation
time \cite{briggs1994effective,chow1997new,kennedy1999partial} and brought these
techniques in the main stream: nowadays \SSA{} based \PRE{} is available in every
industrial compiler.

This paper describes code-motion of identical computations, a technique that
uses the information computed for \PRE{} to detect identical computations but
has a transformation phase whose goal differs from \PRE{}: it removes identical
computations from different branches of execution.  These identical computations
in different branches of execution are not redundant computations at run-time
and the number of run-time computations is not reduced. It is not a redundancy
elimination pass, and thus it has different cost function and heuristics than
\PRE{} or CSE. It is more similar to global code scheduling
\cite{dragonbook,click1995global} in the sense that it will only move
computations. We will now consider an example that explains how code hoisting can
also improve performance.

\subsection{Illustrative Example}
\label{subsec:example}
Code hoisting can reduce the critical path length of execution in out of
order machines. As more instructios are available at the hoisting point, the
hardware has more instructions to reorder. Following example illustrates how
hoisting can improve performance by exposing more ILP.

@@

In this code the two subtractions and the division operations can be executed in
parallel because there are no dependencies among them. So the total number of
cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since $C_{div}$ is
usually much greater than $C_{sub}$ \cite{x86,aarch64}.

@@

In fact there are several advantages of code-motion of this kind:
\begin{itemize}
\item It reduces the code size of the program;
\item It improves function inlining heuristics: functions become cheaper to
  inline when their code size decreases because inliner heuristics in \LLVM{}
  depends on instruction count.
\item It expose more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item It reduces branch misprediction penalty on out-of-order processors with
  speculative execution of branches: by hoisting/sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item It reduces interference/register pressure when appropriate cost-model is
  applied. We have used a cost model in our implementation which hoists/sinks only
  when register pressure would potentially reduce.
\item For SIMD architectures, which execute all branches, it will reduce the
  total number of instructions to be executed.
\item It improves passes that do not work well with branches:
  \begin{itemize}
  \item to improve loop vectorization by reducing a loop with control flow to a
    loop with a single BB, should all the instructions in a conditional get
    hoisted or sinked;
  \item to enable more loop invariant code motion (LICM): as LICM does not
    reason about instructions in the context of loops with conditional branches,
    code-hoisting is needed to move instructions out of conditional expressions
    and expose them to LICM.
  \end{itemize}
\end{itemize}

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using ILP which results in prohibitively high compile time. To the best
of our knowledge we have not found any reference which explored code-motion of
identical computation in as much detail as we have done. A part of our
implementation (aggressive code hoisting) is already merged in LLVM trunk,
however, a general implementation of code-motion of identical expressions is
still missing from GCC and LLVM trunk. The main contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move computations out of branches,
\item a cost model to reduce interference and reduce spills,
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number,
\item experimental evaluation of our implementation in \LLVM{} which combines
  SSA based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce code-size (and improve performance in some
  cases).
\end{itemize}

@@

There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current \LLVM{} implementation of
code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
basic blocks: the instructions of two sibling basic blocks are read in the same
time, and all the instructions of the blocks are hoisted to the common parent
block as long as the compiler is able to prove that the instructions are
equivalent.  This implementation does not allow for an easy extension: first in
terms of compilation time overhead the implementation is quadratic in number of
instructions to bisimulate and second, the equivalence of instructions is
computed by comparing the operands which is neither general nor scalabale.

@@

Rosen \cite{rosen1988global} also introduced moving computations from successors
but using value numbering. Their algorithm iterates on computations of same rank
and move the code with identical computations from the sibling branch to a
common dominator if they are very busy \cite{steven1997advanced}.

@@

Barany \cite{barany2013} presented a global scheduler with ILP formulation with
a goal to minimize register pressure. The results they got were not very
promising. It may be because they only used the scheduler for smaller functions
($<1000$ instructions) also they compiled the benchmarks for ARM-Cortex which is
more resilient to register pressure because it has more registers compared to
say X86.

Shobaki \cite{shobaki2013} also recently presented a combinatorial global
scheduler with reasonable performance improvements. It is possible that both
Barany and Shobaki's implmentation will have similar results when compiled for
same target. Also, both suffer from the same problem, although Shobaki not so
much, of large compile times which is not feasible for industrial compilers like
gcc and LLVM.

@@

The algorithm for code motion uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Control Flow Graph (\CFG{}) and the Dominance (DOM) and Post-Dominance (PDOM)
  relations \cite{dragonbook};
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012}.
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given
  a unique number and the expressions that the compiler can prove to be
  identical are given the same number;
\item Memory \SSA{} \cite{novillo2007memory}: it is a factored use-def chain of
  memory operations that the compiler is able to prove are dependent.
\end{itemize}

The code-motion pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates suitable for code-motion
\item compute a point in the program where it is both legal and profitable to
  move the code,
\item transform the code and update data structures to continue iterative code
  motion.
\end{itemize}

@@

The code shown in Section~\ref{subsec:example} is a reduced example that appears
in a hot loop of a proprietary benchmark.  When the expressions are hoisted from
the conditional clauses, the overall performance of that benchmark improves by
$15\%$ on an out-of-order processor due to increased instruction level
parallelism, and better scheduling of the instructions, accommodating for the
long latency of the division operation.

>>>>

\section{Introduction}
\label{sec:intro}
Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like CSE \cite{dragonbook} work very well on single basic blocks but
when it comes to detect redundancies across basic blocks these techniques fall
short: more complex passes like GCSE and \PRE{} have been designed to handle these
cases based on dataflow analysis \cite{morel1979global}.  At first these
techniques were described in the classical data-flow analysis framework, and
later the use of the \SSA{} representation lowered the cost in terms of compilation
time \cite{briggs1994effective,chow1997new,kennedy1999partial} and brought these
techniques in the main stream: nowadays \SSA{} based \PRE{} is available in every
industrial compiler.

This paper describes code-motion of congruent instructions \cite{briggs1997}, a
technique that uses the information computed for \PRE{} to detect identical
computations but has a transformation phase whose goal differs from \PRE{}: it
removes identical computations from different branches of execution.  These
identical computations in different branches of execution are not redundant
computations at run-time and the number of run-time computations is not
reduced. It is not a redundancy elimination pass, and thus it has different cost
function and heuristics than \PRE{} or CSE. It is more similar to global code
scheduling \cite{dragonbook,click1995global} in the sense that it will only move
computations. Code hoisting can reduce the critical path length of execution in
out of order machines. As more instructios are available at the hoisting point,
the hardware has more instructions to reorder. Following example illustrates how
hoisting can improve performance by exposing more ILP.

@@

In this code the division operation can be executed in parallel with the two
subtractions because there are no dependencies among them. So the total number
of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}.

@@

In fact there are several advantages of code-motion of this kind:
\begin{itemize}
\item It helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because inliner heuristics in \LLVM{} depends on
  instruction count.
\item It expose more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item It reduces branch misprediction penalty on out-of-order processors with
  speculative execution of branches: by hoisting/sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item It reduces interference/register pressure when appropriate cost-model is
  applied. We have used a cost model in our implementation which hoists/sinks only
  when register pressure would potentially reduce.
\item For SIMD architectures, which execute all branches, it will reduce the
  total number of instructions to be executed.
\item It may improve loop vectorization by reducing a loop with control flow to
  a loop with a single BB, should all the instructions in a conditional get
  hoisted or sinked;
\item It enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches the context of loops, code-hoisting is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using ILP which results in prohibitively high compile time. To the best
of our knowledge we have not found any reference which explored code-motion of
identical computation in as much detail as we have done. A part of our
implementation (aggressive code hoisting) is already merged in LLVM trunk,
however, a general implementation of code-motion of congruent (but not
redundant) expressions is still missing from GCC and LLVM trunk. The main
contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructiosn out of branches,
\item a cost model to reduce interference and hence, reduce spills,
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number,
\item experimental evaluation of our implementation in \LLVM{} which combines
  SSA based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce code-size (and improve performance in some
  cases).
\end{itemize}

@@

There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current \LLVM{} implementation of
code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
basic blocks: the instructions of two sibling basic blocks are read at the same
time, and all the instructions of the blocks are hoisted to the common parent
block as long as the compiler is able to prove that the instructions are
equivalent.  This implementation does not allow for an easy extension: first in
terms of compilation time overhead the implementation is quadratic in number of
instructions to bisimulate and second, the equivalence of instructions is
computed by comparing the operands which is neither general nor scalabale.

@@

Rosen \cite{rosen1988global} also briefly discuss hoisting computations with
identical value numbers from immediate successors. Their algorithm iterates on
computations of same rank and move the code with identical computations from the
sibling branch to a common dominator if they are very busy
\cite{steven1997advanced}. The cost-model to mitigate register pressure is
missing, also there is no mention of sinking congruent instructions.

@@

Barany \cite{barany2013} presented a global scheduler with integer linear
programming (ILP) formulation with a goal to minimize register pressure. The
results they got were not very promising. It may be because they only used the
scheduler for smaller functions ($<1000$ instructions) also they compiled the
benchmarks for ARM-Cortex which is more resilient to register pressure because
it has more registers compared to say X86.

Shobaki \cite{shobaki2013} also recently presented a combinatorial global
scheduler with reasonable performance improvements. It is possible that both
Barany and Shobaki's implmentation will have similar results when compiled for
same target architecture. Also, both suffer from the same problem, although
Shobaki not so much, of large compile times which is not feasible for industrial
compilers like gcc and LLVM.

@@

The algorithm for code motion uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
  Control Flow Graph (\CFG{}).
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012}.
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given
  a unique number and the expressions that the compiler can prove to be
  identical are given the same number;
\item Memory \SSA{} \cite{novillo2007memory}: it is a factored use-def chain of
  memory operations that the compiler is able to prove are dependent.
\end{itemize}

The code-motion pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates (congruent instructions) suitable for code-motion,
\item compute a point in the program where it is both legal and profitable to
  move the code,
\item move the code to hoist point or sink point, and
\item update data structures to continue iterative code motion.
\end{itemize}

@@

The code shown in Section~\ref{sec:intro} is a reduced example that appears in a
hot loop of a proprietary benchmark.  When the expressions are hoisted from the
conditional clauses, the overall performance of that benchmark improves by
$15\%$ on an out-of-order processor due to increased instruction level
parallelism, and better scheduling of the instructions, accommodating for the
long latency of the division operation.
